### 2014-11-27 doc/thesis/system.tex
Minor improvements to the text about typing rules

Literal types represent the type of literal values.
They can be the boolean values $\False$ and $\True$,
any integer value, any floating point value, and any string value.
We shall see that literal types are important to represet table types
as records.

@@

We could have omitted the rules \textsc{S-LITERAL}, \textsc{S-BASE},
and \textsc{S-INT2}, because the first two are reflexive and we
could achieve the third one trhough transitivity.
More precisely, we could have defined subsumption rules that
specify the reflexive and trasitive behavior of subtyping,
and use these rules instead of the ones we defined.
For instance, we would not need \textsc{S-INT2} because we could
use the transisitive subsumption rule to derive that
\[
\dfrac{\senv \vdash 1 \subtype \Integer \;\;\;
       \senv \vdash \Integer \subtype \Number}
      {\senv \vdash 1 \subtype \Number}
\]

@@

The rule \textsc{S-TABLE1} introduced the relation $\subtype_{c}$.
It is just an auxiliary relation that defines the subtyping of the
type of the values stored in table fields.
We needed an auxiliary relation because the subtyping of the type
of the values stored in the table fields change according to the
tag of the table type.
For closed tables, subtyping handles mutable fields invariantly,
and imutable fields covariantly.
The following rules define this behavior.
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2} \;\;\;
       \senv \vdash v_{2} \subtype v_{1}}
      {\senv \vdash v_{1} \subtype_{c} v_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash \Const \; v_{1} \subtype_{c} \Const \; v_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash v_{1} \subtype_{c} \Const \; v_{2}}
\end{array}
\end{array}
\]

There are two limitations on \emph{closed} table types that led us to
introduce \emph{open} and \emph{unique} table types.
First, they do not allow programmers to use the table constructor to
initialize table types that include optional fields.
For instance,
\begin{align*}
\{ & ``firstname":\String, ``lastname":\String \}_{closed} \not \subtype\\
\{ & ``firstname":\String,\\
  &  ``middlename":\String \cup \Nil, ``lastname":\String \}_{closed}
\end{align*}
Second, they do not allow programmers to use the table construtor to
initialize table types that represent hashes.
For instance,
\begin{align*}
\{ & ``x":\Number, ``y":\Number, ``z":\Number \}_{closed} \not \subtype\\
\{ & \String:\Number \cup \Nil\}_{closed}
\end{align*}

We introduced \emph{open} table types to allow the refinement
of table types, and also to allow programmers to omit optional
fields in the initialization of table types.
To do that, we defined a subtyping rule that allow \emph{open}
table types to omit fields when $\Nil$ is subtype of the type of
the value of the ommited key.
The following subtyping rule show this behavior:
\[
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
       \senv \vdash v_{i} \subtype_{c} v_{i}' \;\;\;
       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open}}
\end{array}
\]

@@

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = el \; | \;
m \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; el \; | \;
e(el)_{s} \; | \;
e{:}n(el)_{s}\\
e ::= & \;\; \mathbf{nil} \; | \;
c \; | \;
{...}_{e} \; | \;
n \; | \;
e_{1}[e_{2}] \; | \;
{<}t{>} \; n \; | \;
f \; | \;
\{ \} \; | \;
\{ \; \vec{p} \; \} \; | \;
\{ \; {...}_{m} \; \} \; | \;
\{ \; \vec{p},{...}_{m} \; \}\\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
e(el)_{e} \; | \;
e{:}n(el)_{e}\\
l ::= & \;\; n \; | \;
e_{1}[e_{2}] \; | \;
n[c] \; {<}t{>}\\
c ::= & \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string}\\
p ::= & \;\; [e_{1}] = e_{2}\\
el ::= & \;\; \mathbf{nothing} \; | \;
\vec{e} \; | \;
me \; | \;
\vec{e}, me\\
me ::= & \;\; e(el)_{m} \; | \;
e{:}n(el)_{m} \; | \;
{...}_{m}\\
m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; fb\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
f ::= & \;\; \mathbf{fun} \; (){:}r \; fb \; | \;
\mathbf{fun} \; ({...}{:}t){:}r \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t}){:}r \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
fb ::= & \;\; s \;;\; \mathbf{return} \; el\\
n ::= & \;\; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$el$ are expression lists, $c$ are constructor fields, $cl$ are lists of
constructor fields, $me$ are expressions with multiple results,
$f$ are function declarations, $m$ are method declarations, $n$ are variable names,
$t$ are first-level types, and $r$ are second-level types.
The notation $\vec{n{:}t}$ denotes the non-empty list
$n_{1}{:}t_{1}, ..., n_{n}{:}t_{n}$.

@@

We will present the typing rules as a deduction system for the typing
relations $\env_{1} \vdash s : \env_{2}$ and $\env_{1} \vdash e : t, \env_{2}$. 
We will use the first one for typing statements, and the sencond one
for typing expressions.
The first relation means that given the type environment $\env_{1}$,
typing the statement $s$ produces the new type environment $\env_{2}$. 
The sencond relation means that given the type environment $\env_{1}$,
the expression $e$ has type $t$, and produces the new type environment $\env_{2}$.
We need two typing relations because only expressions have types,
though both expressions and statements can modify the environment.
The environment maps variables to types, and we will use $\env_{1}[n \mapsto t]$
for meaning that we extended the environment $\env_{1}$ with the
variable $n$ that maps to type $t$, and this mapping does not affect
the other mappings that extisted before.

@@

Lua has multiple assignments, so Typed Lua also has to take this into
consideration while refining table types.
The rule \textsc{T-ASSINGMENT} shows that tuple types make it easy
to use the consistent-subtyping relation to check multiple assignments:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2} \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1} \vdash \vec{l} = el:\env_{3}}
\end{array}
\end{array}
\]

@@

The next example does not type check because it tries to add
the same field to $a$:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>}, a[``x"] \; {<}\Integer{>} = ``foo", 1$}\\
\end{tabular}
\end{center}

@@

All the examples that we presented until now use the simplest form of
the table constructor: the empty table $\{\}$.
Its inference rule is starightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env \vdash \{\}:\{\}_{unique}, \env
\end{array}
\]

Our abstract syntax reduces the more complex uses of the table
constructor into three forms: $\{\;\vec{p}\;\}$, $\{\;{...}_{m}\;\}$,
and $\{\;\vec{p},{...}_{m}\;\}$.
The first one uses a list of expression pairs $[e_{1}] = e_{2}$,
the second one uses just the vararg expression, and the third one uses both.
We did not include in the abstract syntax a table constructor
that uses a list of expressions because we can write it using the first form.
For instance, we can write the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to mean the table constructor
$\{ ``x", ``y", ``z" \}$.
We chose to not include this form because its inference rule is similar
to the inference rule of the first form, and just controls the type of
the ommited indices.
For the same reason we did not include table constructors with
multiple expressions that can return a tuple type that does not end
with a variadic type.
We did include the vararg expression in the last position to show how
our type system can infer a table type that has a record part and
also an array part.

@@

The inference rules for expression pairs would be straightforward
if our table types allowed any type in the keys.
The rules \textsc{T-PAIR1}, \textsc{T-PAIR2}, and \textsc{T-PAIR3}
check if the type of the key is a $\Boolean$, a $\Number$,
or a $\String$, respectively:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-PAIR1}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Boolean}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-PAIR2}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Number}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-PAIR3}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \String}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
\end{array}
\end{array}
\]

@@

The rule \textsc{T-IDREAD} uses the predicate \emph{close} to do
this promotion before we use a local variable:
\[
\begin{array}{c}
\mylabel{T-IDREAD}\\
\dfrac{\env_{1}(n_{r}) = t}
      {\env_{1} \vdash n_{r}:close(t), \env_{1}}
\end{array}
\]

This rule does not change the type of the local variable because it
just need to ensure that an alias will not get the same type.
For instance, it makes this unsafe example not type check:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{``x":\String \cup \Nil\}_{closed} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String \}_{unique} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$a = b \;;\; a[``x"] = \mathbf{nil}$}
\end{tabular}
\end{center}

@@

We cannot add the field $``x"$ to $b$ because its type is \emph{closed},
and thus do not allow chaning the value that is stored in the field $``x"$
of local $a$.

@@

\[
\begin{array}{c}
\mylabel{T-INVOKE1}\\
\dfrac{\env_{1} \vdash e[n]:p_{1} \rightarrow r, \env_{2} \;\;\;
       \env_{2} \vdash el:p_{2}, \env_{3} \;\;\;
       \Self \times p_{2} \lesssim p_{1}}
      {\env_{1} \vdash e{:}n(el)_{m}:r, \env_{3}}
\end{array}
\]

\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{open} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
\multicolumn{4}{l}{$\mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:number, y:number):\Self$}\\
& & \multicolumn{2}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{closed} =$}\\
& & & \multicolumn{1}{l}{$setmetatable(\{\}, \{ [``\string_\string_index"] = self \})$}\\
& & \multicolumn{2}{l}{$\mathbf{in} \; s[``x"] = x; \; s[``y"] = y; \; \mathbf{return} \; s$}\\
; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:number, y:number):\Void$}\\
& & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y$}
\end{tabular}
\end{center}

@@

@@

The rule \textsc{T-OR1} is the rule that implements the short circuit.
We use the consistent-subtyping in this relation to guarantee that
the type system avaliates the second expression when the first one
has the dynamic type, as it can be hiding a false value.

@@

The rule \textsc{T-OR4} is the most general rule, but it is also
the rule that handles the common $\mathbf{or}$ idioms.
It uses the predicate \emph{filter} to filter possible false values
that might be part of the type of the first expression.
We can use pattern matching to the define the recursive predicate
\emph{filter} as follows:
\begin{align*}
fu(t_{1} \cup t_{2}, t_{3}) & = fu(t_{1}, t_{3}) \cup fu(t_{2}, t_{3})\\
fu(t_{1} \cup t_{2}, t_{1}) & = fu(t_{2}, t_{1})\\
fu(t_{1} \cup t_{2}, t_{2}) & = fu(t_{1}, t_{2})\\
fu(t_{1}, t_{2}) & = t_{1}
\end{align*}

@@

Another common idiom that programmers use in Lua is to overload
the input parameter of functions, and use the function \texttt{type}
to execute different actions according to their types.
\[
\begin{array}{c}
\mylabel{T-IF2}\\
\dfrac{\begin{array}{c}
       closeall(\env_{1}[n \mapsto \String]) \vdash s_{1}:\env_{2} \\
       closeall(\env_{1}[n \mapsto filter(\env_{1}(n), \String)) \vdash s_{2}:\env_{3}
      \end{array}}
      {\env_{1} \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}:closeset(\env_{2}, fav(s_{1}) \cup fav(s_{2}))}

@@

>>>>

Literal types represent the type of literal values.
They can be the boolean values $\False$ and $\True$,
any integer value, any floating point value, and any string value.
We shall see that literal types are important to represent table types
as records.

@@

We could have omitted the rules \textsc{S-LITERAL}, \textsc{S-BASE},
and \textsc{S-INT2}, because the first two are reflexive and we
could achieve the third one through transitivity.
More precisely, we could have defined subsumption rules that
specify the reflexive and transitive behavior of subtyping,
and use these rules instead of the ones we defined.
For instance, we would not need \textsc{S-INT2} because we could
use the transitive subsumption rule to derive that
\[
\dfrac{\senv \vdash 1 \subtype \Integer \;\;\;
       \senv \vdash \Integer \subtype \Number}
      {\senv \vdash 1 \subtype \Number}
\]

@@

The rule \textsc{S-TABLE1} introduced the relation $\subtype_{c}$.
It is just an auxiliary relation that defines the subtyping of the
type of the values stored in table fields.
We needed an auxiliary relation because the subtyping of the type
of the values stored in the table fields change according to the
tag of the table type.
For closed tables, subtyping handles mutable fields invariantly,
and immutable fields covariantly.
The following rules define this behavior.
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FIELD1}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2} \;\;\;
       \senv \vdash v_{2} \subtype v_{1}}
      {\senv \vdash v_{1} \subtype_{c} v_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FIELD2}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash \Const \; v_{1} \subtype_{c} \Const \; v_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FIELD3}\\
\dfrac{\senv \vdash v_{1} \subtype v_{2}}
      {\senv \vdash v_{1} \subtype_{c} \Const \; v_{2}}
\end{array}
\end{array}
\]

There are two limitations on \emph{closed} table types that led us to
introduce \emph{open} and \emph{unique} table types.
First, they do not allow programmers to use the table constructor to
initialize table types that include optional fields.
For instance,
\begin{align*}
\{ & ``firstname":\String, ``lastname":\String \}_{closed} \not \subtype\\
\{ & ``firstname":\String,\\
  &  ``middlename":\String \cup \Nil, ``lastname":\String \}_{closed}
\end{align*}
Second, they do not allow programmers to use the table constructor to
initialize table types that represent hashes.
For instance,
\begin{align*}
\{ & ``x":\Number, ``y":\Number, ``z":\Number \}_{closed} \not \subtype\\
\{ & \String:\Number \cup \Nil\}_{closed}
\end{align*}

We introduced \emph{open} table types to allow the refinement
of table types, and also to allow programmers to omit optional
fields in the initialization of table types.
To do that, we defined a subtyping rule that allow \emph{open}
table types to omit fields when $\Nil$ is subtype of the type of
the value of the omitted key.
The following subtyping rule show this behavior:
\[
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
       \senv \vdash v_{i} \subtype_{c} v_{i}' \;\;\;
       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open}}
\end{array}
\]

@@

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = el \; | \;
m \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; el \; | \;
e(el)_{s} \; | \;
e{:}n(el)_{s}\\
e ::= & \;\; \mathbf{nil} \; | \;
c \; | \;
{...}_{e} \; | \;
n_{e} \; | \;
e_{1}[e_{2}] \; | \;
{<}t{>} \; n \; | \;
f \; | \;
\{ \} \; | \;
\{ \; \vec{p} \; \} \; | \;
\{ \; {...}_{m} \; \} \; | \;
\{ \; \vec{p},{...}_{m} \; \}\\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
e(el)_{e} \; | \;
e{:}n(el)_{e}\\
l ::= & \;\; n_{l} \; | \;
e_{1}[e_{2}] \; | \;
n[c] \; {<}t{>}\\
c ::= & \;\; \mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string}\\
p ::= & \;\; [e_{1}] = e_{2}\\
el ::= & \;\; \mathbf{nothing} \; | \;
\vec{e} \; | \;
me \; | \;
\vec{e}, me\\
me ::= & \;\; e(el)_{m} \; | \;
e{:}n(el)_{m} \; | \;
{...}_{m}\\
m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; fb\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; fb \; | \;
\mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
f ::= & \;\; \mathbf{fun} \; (){:}r \; fb \; | \;
\mathbf{fun} \; ({...}{:}t){:}r \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t}){:}r \; fb \; | \;
\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; fb\\
fb ::= & \;\; s \;;\; \mathbf{return} \; el\\
n ::= & \;\; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$el$ are expression lists, $c$ are literal constants, $p$ are expression pairs,
$me$ are expressions with multiple results, $f$ are function declarations,
$m$ are method declarations, $n$ are variable names,
$t$ are first-level types, and $r$ are second-level types.
The notation $\vec{n{:}t}$ denotes the non-empty list
$n_{1}{:}t_{1}, ..., n_{n}{:}t_{n}$.

@@

We also included two kinds of type cast in our core language:
the expression ${<}t{>} \;n$ and the left-hand value $n[c] \; {<}t{>}$.
We will show that the first helps allowing safe aliasing between
table types, while the second allows the refinement of table types.
Due to the refinement of table types we also split variable names
into two categories: $n_{e}$ when they appear as expressions and
$n_{l}$ when they appear as left-hand values.

We will present the typing rules as a deduction system for the typing
relations $\env_{1} \vdash s : \env_{2}$ and $\env_{1} \vdash e : t, \env_{2}$. 
We will use the first one for typing statements, and the second one
for typing expressions.
The first relation means that given the type environment $\env_{1}$,
typing the statement $s$ produces the new type environment $\env_{2}$. 
The second relation means that given the type environment $\env_{1}$,
the expression $e$ has type $t$, and produces the new type environment $\env_{2}$.
We need two typing relations because only expressions have types,
though both expressions and statements can modify the environment.
The environment maps variables to types, and we will use $\env_{1}[n \mapsto t]$
for meaning that we extended the environment $\env_{1}$ with the
variable $n$ that maps to type $t$, and this mapping does not affect
the other mappings that existed before.

@@

Lua has multiple assignments, so Typed Lua also has to take this into
consideration while refining table types.
The rule \textsc{T-ASSINGMENT} shows that tuple types make it easy
to use the consistent-subtyping relation to type check multiple assignments:
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2} \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1} \vdash \vec{l} = el:\env_{3}}
\end{array}
\end{array}
\]

@@

The next example does not type check because it tries to add
the same field to $a$, but with different types:
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$a[``x"] \; {<}\String{>}, a[``x"] \; {<}\Integer{>} = ``foo", 1$}\\
\end{tabular}
\end{center}

@@

All the examples that we presented until now use the simplest form of
the table constructor: the empty table $\{\}$.
Its inference rule is straightforward:
\[
\begin{array}{c}
\mylabel{T-CONSTRUCTOR1}\\
\env \vdash \{\}:\{\}_{unique}, \env
\end{array}
\]

Our abstract syntax reduces the more complex uses of the table
constructor into three forms: $\{\;\vec{p}\;\}$, $\{\;{...}_{m}\;\}$,
and $\{\;\vec{p},{...}_{m}\;\}$.
The first one uses a list of expression pairs $[e_{1}] = e_{2}$,
the second one uses just the vararg expression, and the third one uses both.
We did not include in the abstract syntax a table constructor
that uses a list of expressions because we can write it using the first form.
For instance, we can write the table constructor
$\{ [1] = ``x", [2] = ``y", [3] = ``z" \}$ to express $\{ ``x", ``y", ``z" \}$.
We chose to not include this form because its inference rule is similar
to the inference rule of the first form, and just controls the type of
the omitted indices.
For the same reason we did not include table constructors with
multiple expressions, which can return a tuple type that does not end
with a variadic type.
We did include the vararg expression in the last position to show how
our type system can infer a table type that has a record part and
also an array part.

@@

The inference rules for expression pairs would be straightforward
if our table types allowed any type in the keys.
The rules \textsc{T-PAIR1}, \textsc{T-PAIR2}, and \textsc{T-PAIR3}
check if the type of the key is a subtype of $\Boolean$, $\Number$,
or $\String$, respectively:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-PAIR1}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Boolean}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-PAIR2}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \Number}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-PAIR3}\\
\dfrac{\env_{1} \vdash e_{1}:t_{1}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t_{2}, \env_{3} \;\;\;
       t_{1} \subtype \String}
      {\env_{1} \vdash [e_{1}] = e_{2}: t_{1}{:}t_{2}, \env_{3}}
\end{array}
\end{array}
\]

@@

The rule \textsc{T-IDREAD} uses the predicate \emph{close} to do
this promotion before we use a local variable:
\[
\begin{array}{c}
\mylabel{T-IDREAD}\\
\dfrac{\env_{1}(n) = t}
      {\env_{1} \vdash n_{e}:close(t), \env_{1}}
\end{array}
\]

This rule does not change the type of the local variable because it
just need to ensure that an alias will not get the same type.
For instance, it makes this unsafe example not type check:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{``x":\String \cup \Nil\}_{closed} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``x":\String \}_{unique} = \{ [``x"] = ``foo" \} \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$a = b \;;\; a[``x"] = \mathbf{nil} \;;\; b[``x"] = b[``x"] \;{..}\; ``bar"$}
\end{tabular}
\end{center}

@@

We cannot add the field $``x"$ to $b$ because its type is \emph{closed},
and thus do not allow changing the value that is stored in the field $``x"$
of local $a$.

@@

The rule \textsc{T-METHOD3} allow our type system to type check the
following example:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{open} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
\multicolumn{4}{l}{$\mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:\Number, y:\Number):\Self$}\\
& & \multicolumn{2}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{closed} =$}\\
& & & \multicolumn{1}{l}{$setmetatable(\{\}, \{ [``\string_\string_index"] = self \})$}\\
& & \multicolumn{2}{l}{$\mathbf{in} \; s[``x"] = x; \; s[``y"] = y; \; \mathbf{return} \; s$}\\
; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:\Number, y:\Number):\Void$}\\
& & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y$}
\end{tabular}
\end{center}

This example uses the type $\{``x":\Number, ``y":\Number\}_{open}$ to
initialize the local variable $s$, which represents the type of the class
that we are defining.
Then, we use two method definitions to include $new$ and $move$ into
our class that now has type
\begin{align*}
\{ & ``x":\Number, ``y":\Number,\\
   & ``new":\Self \times \Number \times \Number \rightarrow \Self,\\
   & ``move":\Self \times \Number \times \Number \rightarrow \Void \}_{open}
\end{align*}

Inside the definition of the method \emph{new} we use \emph{setmetatable}
to initialize the local $s$ with the type of $\Self$.
The rule \textsc{T-SETMETATABLE} express this idea:
\[
\begin{array}{c}
\mylabel{T-SETMETATABLE}\\
\dfrac{\env_{1} \vdash e:t, \env_{2} \;\;\;
       t \subtype \{\}_{open}}
      {\env_{1} \vdash setmetatable(\{\}, \{[``\string_\string_index"] = e\}):close(t), \env_{2}}
\end{array}
\]

After we define our class, we can use it to create object instances
of this class and call its methods.
The next example assumes the object $o$ is in the environment as
has the type of the class we just defined:
\begin{center}
\begin{tabular}{l}
$o{:}move(10, 10)_{s}$
\end{tabular}
\end{center}

This method call type checks through the rule \textsc{T-INVOKESTM1}.
This rule uses the inference rule \textsc{T-INVOKE1} and discards the results.
We defined the rule \textsc{T-INVOKE1} as follows:
\[
\begin{array}{c}
\mylabel{T-INVOKE1}\\
\dfrac{\env_{1} \vdash e:t, \env_{2} \;\;\;
       \env_{2} \vdash e[n]:p_{1} \rightarrow r, \env_{3} \;\;\;
       \env_{3} \vdash el:p_{2}, \env_{4} \;\;\;
       \Self \times p_{2} \lesssim p_{1}}
      {\env_{1} \vdash e{:}n(el)_{m}:[\Self \mapsto t]r, \env_{4}}
\end{array}
\]

@@

This example does not type check because we are trying to call
$s$ with $\Integer \times \Integer \times \String$,
but its input parameter has type $\Integer \times \Integer$.
Lua drops extra values in function calls, and Typed Lua can also
have this behavior.
If we change the input type of $s$ to $\Integer \times \Integer \times \Value{*}$,
this example type checks.
Typed Lua also has to adjust function calls to use $\Nil$ in the
place of missing parameters, otherwise it would not be able to
type check function calls that have optional parameters.

\textit{/* I guess we should have different typing rules for strict and default. */}

@@

The rule \textsc{T-OR1} is the rule that implements the short circuit.
We use the consistent-subtyping relation in this rule to guarantee that
the type system checks the second expression when the first one
has the dynamic type, as it can be hiding a false value.

@@

The rule \textsc{T-OR4} is the most general rule, but it is also
the rule that handles the common $\mathbf{or}$ idioms.
It uses the predicate \emph{filter} to filter possible false values
that might be part of the type of the first expression.
We can use pattern matching to the define the recursive predicate
\emph{filter} as follows:
\begin{align*}
filter(t_{1} \cup t_{2}, t_{1}) & = filter(t_{2}, t_{1})\\
filter(t_{1} \cup t_{2}, t_{2}) & = filter(t_{1}, t_{2})\\
filter(t_{1} \cup t_{2}, t_{3}) & = filter(t_{1}, t_{3}) \cup filter(t_{2}, t_{3})\\
filter(t_{1}, t_{2}) & = t_{1}
\end{align*}

@@

Another common idiom that programmers use in Lua is to overload
the input parameter of functions, and use the function \texttt{type}
to execute different actions according to their types.
\[
\begin{array}{c}
\mylabel{T-IF2}\\
\dfrac{\begin{array}{c}
       \env_{1}(n) = t\\
       closeall(\env_{1}[n \mapsto \String]) \vdash s_{1}:\env_{2} \\
       closeall(\env_{1}[n \mapsto filter(t, \String)) \vdash s_{2}:\env_{3}
      \end{array}}
      {\env_{1} \vdash \mathbf{if} \; type(n) == ``string" \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}:closeset(\env_{1}[n \mapsto t], fav(s_{1}) \cup fav(s_{2}))}

@@

Although this last example is not safe, it shows that optional
type systems still preserve the flexibility of dynamically
typed languages along with the benefits of static type checking.
